%---------------------------------------------------------------
% Main thesis text
%---------------------------------------------------------------

%---------------------------------------------------------------
\chapter{Introduction}
%---------------------------------------------------------------
\setcounter{page}{1}

\begin{chapterabstract}
This thesis addresses the engineering problem of deploying and evaluating
modern face-processing neural networks directly on mobile devices. The work
focuses on the theoretical foundations and experimental methodology required
for trustworthy benchmarking of face detection and facial landmark estimation
on Android and iOS using ExecuTorch and backend delegates.
\end{chapterabstract}

\section{Motivation}

On-device inference has become a key requirement for many user-facing
applications, because it improves privacy, reduces round-trip latency, and
allows offline operation. At the same time, mobile deployment is constrained by
strict limits on compute throughput, memory bandwidth, power, and thermal
budget \cite{wang2020mobilesoc,li2024mobilebenchmark}. In practice, this
means that achieving good model accuracy is not sufficient; the full
end-to-end system must also be efficient and stable under realistic operating
conditions.

Recent community benchmarks confirm that measured latency strongly depends on
hardware configuration, runtime backend, numerical precision, and measurement
protocol \cite{reddi2022mlperfmobile,li2024mobilebenchmark}. Therefore,
reliable conclusions require not only model comparison but also careful
benchmark design and reproducible data processing.

\section{Problem statement}

The core problem of this thesis is to design a methodologically sound framework
for evaluating mobile face-processing pipelines implemented with ExecuTorch.
The framework must support equivalent iOS and Android applications, consistent
measurement conditions, and post-hoc validation of quality metrics.

The benchmarked workload is an end-to-end two-stage pipeline:
\begin{enumerate}
    \item face detection in an input image,
    \item facial landmark estimation on detected face crops.
\end{enumerate}

The thesis intentionally focuses on theoretical and methodological parts.
Numerical benchmark outcomes are out of scope of this text.

\section{Research objectives}

Theoretical and engineering objectives of the thesis are:
\begin{itemize}
    \item summarize the main concepts of on-device inference relevant to mobile
          face-processing;
    \item describe ExecuTorch architecture and deployment workflow for mobile
          targets \cite{executorch_how,executorch_arch};
    \item define a reproducible benchmark protocol for latency, throughput, and
          landmark accuracy;
    \item identify validity threats (thermal throttling, backend asymmetry,
          unsupported operators, logging artifacts) and propose mitigations.
\end{itemize}

\section{Main contributions}

The thesis contributes:
\begin{itemize}
    \item a structured theoretical overview connecting mobile hardware
          constraints, model optimization, and runtime system design;
    \item an ExecuTorch-focused deployment explanation covering export,
          lowering, delegation, and runtime execution
          \cite{executorch_export,executorch_runtime,executorch_delegate};
    \item a practical methodology for fair cross-platform benchmarking of
          detector--landmark pipelines;
    \item a reproducibility-oriented data handling workflow suitable for
          independent audit and future extension.
\end{itemize}

\section{Thesis structure}

Chapter~2 introduces theoretical background: mobile inference constraints,
efficient model design, quantization, and face-processing tasks. Chapter~3
explains ExecuTorch architecture and backend delegation for mobile deployment.
Chapter~4 defines the benchmark methodology, metrics, protocol, and validity
controls. The final chapter concludes the thesis.

%---------------------------------------------------------------
\chapter{Theoretical background}
%---------------------------------------------------------------

\section{On-device inference in mobile environments}

Mobile AI systems are typically deployed on heterogeneous Systems-on-Chip
(SoCs), where CPU, GPU, and dedicated NPUs coexist
\cite{lee2019mobilegpu,wang2020mobilesoc}. Compared to server inference,
mobile workloads operate under tighter thermal and power limits and often share
resources with interactive foreground tasks.

From a systems perspective, latency depends on several layers:
\begin{itemize}
    \item model architecture (operator types, tensor shapes, parameter count),
    \item runtime backend (kernel library, operator fusion, threading policy),
    \item hardware mapping (CPU/GPU/NPU dispatch),
    \item memory subsystem behavior (cache locality, bandwidth pressure),
    \item thermal governor and dynamic voltage/frequency scaling (DVFS).
\end{itemize}

As a consequence, the same neural network can show substantially different
performance across devices and software stacks
\cite{li2024mobilebenchmark,reddi2022mlperfmobile}.

\section{Hardware and system constraints}

\subsection{CPU, GPU, and NPU roles}

CPUs provide broad operator coverage and predictable behavior, but may not offer
the highest throughput for dense tensor computations. Mobile GPUs can improve
throughput for selected workloads, while NPUs may deliver superior
energy-efficiency when supported operators map well to accelerator kernels
\cite{lee2019mobilegpu,wang2020mobilesoc}. In real deployments, unsupported
subgraphs frequently fall back to CPU, introducing synchronization and memory
transfer overhead.

\subsection{Power and thermal behavior}

Peak performance is often not sustainable in handheld devices. Prolonged
inference can trigger thermal throttling, reducing frequency and increasing
latency variance. Benchmark methodology must therefore control ambient
conditions, warm-up state, and per-run timing windows
\cite{li2024mobilebenchmark}.

\section{Efficient neural networks for mobile vision}

The shift from server-centric CNNs to mobile-friendly architectures was enabled
by operator-level efficiency ideas such as depthwise separable convolution,
inverted residual blocks, and channel shuffle
\cite{howard2017mobilenets,sandler2018mobilenetv2,zhang2018shufflenet}.
SqueezeNet demonstrated that competitive accuracy can be achieved with
significantly fewer parameters \cite{iandola2016squeezenet}.

For mobile face-processing, both detector and landmark models must be efficient.
The detector controls how many candidate faces are forwarded to the second
stage, while the landmark model dominates per-face regression cost.

\section{Model compression and quantization}

Compression and quantization reduce memory footprint and often improve runtime
performance. Classical compression includes pruning and weight sharing
\cite{han2016deepcompression}. In practical mobile deployment, integer
quantization is especially important because INT8 kernels can be faster and more
energy-efficient than FP32 on supported hardware \cite{jacob2018quantization}.

Post-training quantization (PTQ) is attractive for deployment pipelines because
it does not require full retraining. However, sensitivity varies across models
and layers; accuracy degradation can occur without representative calibration
and proper backend support \cite{krishnamoorthi2018quantizing}.

\section{Face detection and landmark estimation}

\subsection{Task definitions}

Face detection localizes faces with bounding boxes. Landmark estimation predicts
keypoint coordinates (e.g., eye corners, nose tip, mouth corners), usually on a
cropped face image. These tasks are strongly coupled in end-to-end systems:
detection quality and crop geometry directly influence landmark quality.

\subsection{Representative models}

RetinaFace is a strong one-stage detector with high robustness under challenging
conditions \cite{deng2020retinaface}. FaceBoxes is a lightweight detector
explicitly designed for CPU real-time operation \cite{zhang2017faceboxes}.

For landmarks, PFLD emphasizes compactness and speed \cite{guo2019pfld}.
MobileFaceNet provides a lightweight backbone originally developed for mobile
face recognition and often reused for landmark regression
\cite{chen2018mobilefacenets}. PIPNet combines heatmap-style reasoning with
low-resolution efficient heads for practical landmark localization
\cite{jin2021pipnet}.

\subsection{Datasets and annotation uncertainty}

In this thesis, AFW annotations are used for wild-condition landmark evaluation,
originating from the face detection and pose estimation work of Zhu and
Ramanan \cite{zhu2012facedetectionwild}. As with most in-the-wild datasets,
annotation noise, partial occlusion, and pose diversity should be treated as
intrinsic uncertainty rather than pure model error.

\section{Benchmarking principles}

Prior benchmarking studies stress that meaningful comparisons require controlled
experimental factors: fixed run rules, warm-up handling, hardware affinity,
consistent data preprocessing, and transparent logging
\cite{reddi2022mlperfmobile,li2024mobilebenchmark}. Consequently, this thesis
adopts a protocol-first approach: theory and methodology are defined before any
result interpretation.

%---------------------------------------------------------------
\chapter{ExecuTorch for mobile deployment}
%---------------------------------------------------------------

\section{Why ExecuTorch}

ExecuTorch is a PyTorch edge deployment stack designed for efficient on-device
inference across mobile and embedded targets
\cite{executorch_how,executorch_arch}. Its central design principle is
\emph{ahead-of-time} (AOT) preparation: expensive transformation and planning
steps are performed before deployment, so runtime execution remains lightweight.

\section{Conceptual architecture}

ExecuTorch organizes model deployment into three phases
\cite{executorch_how,executorch_arch}:
\begin{enumerate}
    \item export a PyTorch model to graph form,
    \item lower and optimize the graph for edge execution,
    \item execute the serialized program with a small C++ runtime.
\end{enumerate}

The terminology of ATen/Core ATen/Edge/Backend dialects defines progressive
lowering from general PyTorch representation toward target-specific execution
\cite{executorch_concepts}.

\section{Export and lowering workflow}

The practical workflow consists of:
\begin{enumerate}
    \item model export using \texttt{torch.export},
    \item backend-aware lowering via
          \texttt{to\_edge\_transform\_and\_lower},
    \item emission of a serialized \texttt{.pte} artifact.
\end{enumerate}

This process supports dynamic shape constraints and optional separation of
program and tensor data into \texttt{.pte} and \texttt{.ptd} files
\cite{executorch_export,executorch_pte,executorch_ptd,pytorch_export}.

\section{Delegation and partitioning}

ExecuTorch uses delegates to offload supported subgraphs to backend-specific
implementations \cite{executorch_delegate}. A partitioner tags lowerable graph
regions; unsupported regions remain in portable execution. This mixed execution
model improves portability but can create overhead from frequent backend
boundaries.

For this thesis, two backends are central:
\begin{itemize}
    \item \textbf{XNNPACK}: optimized CPU backend with broad mobile support,
          including quantized paths \cite{executorch_xnnpack,xnnpack_repo};
    \item \textbf{Core ML delegate}: Apple backend path that can dispatch to CPU,
          GPU, or ANE depending on model and platform capabilities
          \cite{executorch_coreml,apple_coreml}.
\end{itemize}

\section{Runtime design and implications for benchmarking}

The runtime loads \texttt{.pte} programs, resolves operator kernels or delegate
entry points, and executes a preplanned instruction sequence
\cite{executorch_runtime}. Because planning is mostly AOT, runtime overhead is
low and easier to reason about than dynamic graph execution.

For benchmarking, this has three practical implications:
\begin{itemize}
    \item model initialization and first-run costs must be separated from
          steady-state latency;
    \item backend registration/linking correctness is mandatory, otherwise runs
          silently degrade to slower fallback paths;
    \item exported artifact provenance (backend, precision, shape constraints)
          must be tracked in run metadata.
\end{itemize}

\section{Quantization in ExecuTorch}

ExecuTorch integrates quantization workflows through backend-specific
configurations, commonly using PT2E/torchao-based preparation and conversion
\cite{executorch_quantization,torchao_pt2e}. Since quantization support differs
across delegates and operators, each quantized artifact must be validated for
both numerical behavior and actual backend coverage.

\section{Practical limitations}

In real projects, deployment quality is affected by operator support matrices,
toolchain version compatibility, and backend-specific converter behavior. These
issues are not edge cases; they are a normal part of mobile ML engineering and
must be documented alongside benchmark methodology.

%---------------------------------------------------------------
\chapter{Benchmark methodology}
%---------------------------------------------------------------

\section{Methodological goals}

The benchmark design in this thesis follows two goals:
\begin{enumerate}
    \item produce comparable latency/throughput/accuracy measurements across
          iOS and Android,
    \item preserve raw artifacts needed for independent recomputation and
          audit.
\end{enumerate}

The approach extends prior benchmarking ideas from mobile inference literature
\cite{li2024mobilebenchmark,reddi2022mlperfmobile} and adapts them to
face-processing pipelines.

\section{Evaluated pipeline}

Each benchmark run executes an end-to-end pipeline:
\begin{enumerate}
    \item detector preprocessing,
    \item detector inference,
    \item detector postprocessing (box decoding and filtering),
    \item face crop generation,
    \item landmark model preprocessing,
    \item landmark inference,
    \item landmark postprocessing.
\end{enumerate}

End-to-end timing includes all stages above, not only neural forward passes.
This reflects application-level latency perceived by users.

\section{Model families and configuration space}

The methodology supports multiple detector/landmark pairs and backend
configurations. Typical factors include:
\begin{itemize}
    \item detector choice (e.g., RetinaFace vs FaceBoxes),
    \item landmark architecture (e.g., PFLD, MobileFaceNet, PIPNet),
    \item precision variant (FP32, FP16 where available, INT8),
    \item backend delegate (XNNPACK, Core ML, or portable fallback),
    \item resolved batch size.
\end{itemize}

A complete run record stores these factors explicitly, so later analysis does
not depend on file naming conventions alone.

\section{iOS and Android implementation parity}

Because platform asymmetries can bias conclusions, iOS and Android benchmark
applications are designed to be feature-equivalent at the pipeline level.
Equivalent behavior includes:
\begin{itemize}
    \item identical preprocessing parameters,
    \item identical postprocessing logic and thresholds,
    \item consistent metric collection and export schema,
    \item consistent handling of warm-up and measured iterations.
\end{itemize}

This principle follows the project guideline that iOS and Android applications
should remain equivalent in features and purpose.

\section{Dataset and inputs}

AFW-based landmark annotations are used for accuracy-oriented offline
recomputation \cite{zhu2012facedetectionwild}. Input assets and ordering are
versioned as part of run metadata. For each run, used-input logs are exported to
avoid ambiguity caused by dynamic file-system ordering.

\section{Metrics}

\subsection{Latency and throughput}

Let $T_{\text{run}}$ be measured end-to-end runtime for $N$ processed frames. The
average throughput is
\begin{equation}
\mathrm{FPS} = \frac{N}{T_{\text{run}}}.
\end{equation}

Per-frame latency is measured in milliseconds and summarized by robust
statistics (median and selected percentiles), not only arithmetic means.

\subsection{Landmark accuracy}

For $N$ faces and $K$ landmarks, normalized mean error (NME) is defined as
\begin{equation}
\mathrm{NME} = \frac{1}{N}\sum_{n=1}^{N}\frac{1}{K}\sum_{k=1}^{K}
\frac{\lVert \hat{\mathbf{p}}_{n,k} - \mathbf{p}_{n,k} \rVert_2}{d_n},
\end{equation}
where $\hat{\mathbf{p}}_{n,k}$ is predicted landmark position,
$\mathbf{p}_{n,k}$ is ground truth, and $d_n$ is the normalization distance
(e.g., inter-ocular distance).

\subsection{Optional energy metrics}

If platform tooling allows reliable collection, energy-related signals can be
logged as supplementary indicators. In this thesis, latency and accuracy metrics
remain primary.

\section{Measurement protocol}

To reduce noise and measurement bias, each benchmark condition follows a fixed
protocol:
\begin{itemize}
    \item dedicated warm-up iterations before timed measurement,
    \item fixed number of measured iterations,
    \item stable device state (charging policy, background app control,
          thermal preconditioning),
    \item separate logging for initialization, inference, and optional
          postprocessing stages.
\end{itemize}

Runs that violate schema, contain missing key payloads, or fail consistency
checks are flagged for exclusion from aggregate statistics.

\section{Run artifacts and offline analysis}

Every run exports a machine-readable summary plus raw samples. Offline analysis
scripts recompute aggregate statistics and, when possible, recompute accuracy
from per-image predictions instead of trusting prefilled summary fields. This
post-hoc recomputation improves robustness against stale metadata and partial
export errors.

\section{Threats to validity}

Main threats and mitigations include:
\begin{itemize}
    \item \textbf{Thermal drift}: controlled warm-up and repeated runs;
    \item \textbf{Backend asymmetry}: shared-configuration comparisons for
          cross-device claims;
    \item \textbf{Unsupported operators}: explicit reporting of delegate
          coverage and fallback paths;
    \item \textbf{Data leakage or ordering artifacts}: deterministic input lists
          and exported input manifests;
    \item \textbf{Metric inconsistency}: central offline recomputation of
          statistics.
\end{itemize}

\section{Ethical and privacy considerations}

The thesis evaluates technical performance of face-processing models and does
not develop identity recognition functionality. Nevertheless, any deployment of
face analytics should follow legal and ethical constraints for biometric data,
including transparent purpose limitation, secure storage, and user consent where
required.

%---------------------------------------------------------------
\chapter{Conclusion}
%---------------------------------------------------------------

This thesis presents the theoretical and methodological basis for benchmarking
mobile face-processing inference with ExecuTorch. The text first establishes
mobile inference constraints, then connects them to efficient model design,
quantization, and runtime backend behavior. It further explains ExecuTorch’s
export-to-runtime path, delegate mechanism, and artifact formats relevant to
practical deployment.

A reproducible benchmark methodology is defined for end-to-end detector plus
landmark pipelines on iOS and Android, with explicit metric definitions,
protocol controls, and validity safeguards. The methodology is designed to
support fair cross-platform comparison and independent recomputation from raw
artifacts.

The resulting framework provides a solid base for empirical evaluation and
future extension, including additional devices, backends, and optimization
strategies. Numerical evaluation outcomes are intentionally outside the scope of
this text.
